### 2018-06-07 ident/docs/draft_4.tex
new draft with corrections from WRC

	Constraint-based models (CBMs) of metabolism are used to improve our understanding of metabolism by representing it as a stoichiometric network of reactions that operate under a pseudo-steady state assumption \parencite{Bordbar2014a}. The ability of CBMs to shine light on the nonintuitive interactions that govern cellular metabolism is leveraged to engineer and asses the impact of designs that alter the ability of a cell to grow, or produce a desired metabolite \parencite{Maia2016}. However, in CBMs, metabolism is assumed to operate under a pseudo steady state. Consequently, the metabolite concentrations within the metabolic network are assumed to be constant, and changes in metabolite concentrations are not modeled. Furthermore, since CBMs represent metabolism using only the stoichiometry of its constituent reactions, they do not account for the various non-catalytic regulatory interactions that are also responsible for metabolic function. These shortcomings prevent CBMs from being used to fully understand the steady state as well as the dynamic characteristics of metabolic networks.

	In contrast, the effects of regulatory interactions and changes in metabolite concentrations on different characteristics of metabolism can be studied using kinetic models of metabolism \parencite{Saa2017}. These models account for changes in metabolite concentrations subject to thermodynamic and regulatory constraints that underly metabolic networks in addition to its stoichiometry \parencite{Link2014}. Kinetic models can not only help us better understand lesser known and understood characteristics of metabolism like bistability \parencite{Kotte2014}, and their role in human health, but can also improve predictions about the impact of engineering design perturbations on metabolism, and help propose alternative designs to achieve metabolite production goals \parencite{Khodayari2016}.

	Kinetic models differ from CBMs in their use of mechanistic enzyme kinetic rate laws to model enzyme catalyzed fluxes within a metabolic network \parencite{Srinivasan2015,Saa2017}. The use of kinetic models requires information on the enzyme kinetic rate laws that will be used to model the fluxes within a metabolic network, as well as numerical values for the parameters used in these rate laws. Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where kinetic rate laws and parameter values play a crucial role \parencite{Srinivasan2017}.

	Despite their importance, the parameterization of kinetic models is still a problem for which solutions are a subject of debate within the modeling community. Typically, enzyme kinetic rate laws are parameterized based on in vitro observations of enzyme activity, as opposed to observations made under in vivo conditions \parencite{Heijnen2005,Smallbone2007}. However, some researchers have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions, as opposed to in vitro conditions \parencite{Heijnen2005,Heijnen2013}. On the other hand, some reports have shown that despite the large uncertainties associated with parameters estimated based on in vivo experimental data \parencite{Link2014}, in vitro parameter estimates are a reasonable approximation of values that would be applicable under in vivo conditions \textcolor{red}{(Ron Milo paper comparing in vitro vs invivo enzyme turn over rates in PLoS Computational Biology)}.

	 Authors have sought to constrain and determine uncertainty in parameter estimates and consequent model predictions by using Monte Carlo approaches for kinetic modeling of metabolism \parencite{Andreozzi2016a}. These approaches also allow for the integration of experimentally observed in vivo data. ORACLE \textcolor{red}{(Wang and Haztzimanikatis, 2004)} and Ensemble modeling \textcolor{red}{(Tan and Liao, 2008)}, are two such examples. These and other Monte Carlo kinetic modeling methods have been previously reviewed \parencite{Srinivasan2015}. 	 
	 Bayesian approaches to improve parameter estimation and quantify estimation uncertainty have also been proposed \parencite{Saa2016}. \cite{Vanlier2013} provide a review of different Bayesian approaches to quantify parameter uncertainty.

	Related to the development of these methods to quantify parameter estimation uncertainty, the importance of model parameter identifiability, i.e., the condition that needs to be satisfied to estimate unique kinetic parameter values from experimental data, is often overlooked \parencite{Ljung1994,Berthoumieux2013}. In a model, a parameter is said to be structurally or a priori identifiable if its values can be uniquely estimated from available experimental data, independent of all other model parameters. However, if a parameter cannot be uniquely estimated due to redundant model parameterization, or due to the nonlinear relationships between the model parameters, then the parameter is said to be structurally non-identifiable. Conversely, if the ability to estimate unique parameter values is compromised due to the inability of the available data to capture the requisite information needed to estimate the parameters in the modeled system, and the uncertainty in parameter estimates is unquantifiable, the parameter is said to be practically non-identifiable \parencite{Ljung1994}.

	Authors have proposed ways to assess parameter identifiability by proposing approximate kinetic models of metabolism that utilize empirical enzyme kinetic rate laws with parameters that have physical significance, and are identifiable \parencite{Heijnen2005,Smallbone2007}. Significant work has also been done towards the development of methods for structural identification of parameters in kinetic models of metabolism \parencite{Ljung1994,Nikerel2009,Berthoumieux2013,Raue2014}\textcolor{red}{(paper from Rudiyanto Gunawan on model discrimination and sensitivity analysis)}.

	Methods to improve practical identifiability through a priori experimental design have also been developed, with a focus on kinetic models of metabolism \parencite{Gadkar2005a,Vanlier2014a,Raue2014}. Some of these methods are limited by their applicability to approximate kinetic models only \parencite{Nikerel2009,Berthoumieux2013}, while some of them suffer from computational limitations when applied to kinetic models of large metabolic networks \parencite{Gadkar2005a,Raue2014}\textcolor{red}{(Banga method using FIM for D-optimal design, ??)}.

	In this paper, we propose a scalable methodology that uses available steady state fluxomics, metabolomics and proteomics data to test the practical identifiability of parameters for each individual reaction in kinetic models of metabolism. We demonstrate how the computer algebra-based method that we have developed to perform this test can also facilitate the design of experiments for generating the data required to estimate unique parameter values for all reaction fluxes in a metabolic network. In doing so, we are able to propose the number and types of perturbations that will provide the most useful data for parameter estimation, as well as test the identifiability of different enzyme kinetic rate laws that are typically used to model fluxes in metabolic networks. We illustrate our methodology using a small metabolic network model of glucoeneogenesis in \textit{Escherichia coli} \parencite{Kotte2014, Srinivasan2017} under the assumption that all intracellular metabolite concentrations and fluxes can be measured.

@@

	Parameter estimation methods based on optimization principles are used to determine true parameter values from experimental data. Under the assumption that all intracellular metabolite concentrations and fluxes can be measured, a parameter estimation problem can be formulated as a nonlinear programming problem (Equation \ref{eq:lsqopt}) to estimate the values of enzyme kinetic parameters, $\theta \in \mathbb{R}^{n_p}$, based on the measured data. 
	\begin{subequations}\label{eq:lsqopt}
		\begin{align}
		\underset{\theta}{\mathrm{min}} &\sum_{k=1}^{n_x + n_r}\sum_{l=1}^{n_t}\left(\frac{y_{kl}^*-y_{kl}}{\sigma_{kl}^*}\right)^2\\
		&\theta_l \le \theta \le \theta_u
		\end{align}
	\end{subequations}
	Here $y \in \mathbb{R}^{n_x + n_r}$ is the combined vector of concentrations ($x\in \mathbb{R}^{n_x}$) and fluxes ($v \in \mathbb{R}^{n_r}$), at each time point $l={1, 2, ..., n_t}$. The minimization of the least squares error between the measured ($y^*$) and modeled ($y$) concentrations and fluxes, weighted by the variance in the experimental data $\sigma_{kl}^*$ for each concentration and flux, at each time point $l={1, 2, ..., n_t}$, is used as an objective function (Equation \ref{eq:lsqopt}a) for the optimization problem. The least squares parameter values are determined within fixed upper ($\theta_u$) and lower ($\theta_l$) bounds (Equation \ref{eq:lsqopt}b).

@@

	The parameters in $\theta$ in any nonlinear model (Equation \ref{eq:kinstoich}) are said to be structurally identifiable if, for an input-output mapping defined by $y = \Phi(\theta,u)$ for at least one input function $u$, any two values of parameters $\theta_1$ and $\theta_2$ satisfy the relationship in Equation (\ref{eq:stident}):
	\begin{align}\label{eq:stident}
	\Phi(\theta_1,u) = \Phi(\theta_2,u) \iff \theta_1 = \theta_2
	\end{align}
	Accordingly, if parameters in $\theta$ have a unique value, a finite number of non-unique values or an infinite number of values for all input functions, they are said to be structurally globally identifiable, locally identifiable or non-identifiable, respectively. So, the structural identifiability of parameters in a dynamic model helps establish the presence or absence of a relationship between the unmeasured and measured concentrations/fluxes, as well as correlations between different model parameters \textcolor{red}{(Rudiyanto Gunawan paper on model discrimination)}. Consequently, the effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter.

	Experimental data from many physical systems is usually noisy, and when parameters are estimated on the basis of noisy data, the ability to estimate unique parameter values to satisfy Equation (\ref{eq:stident}) is referred to as practical identifiability. If a single unique parameter satisfying Equation (\ref{eq:stident}) can be found, then $\theta$ is said to be globally practically identifiable. Whereas, if parameter estimates with quantifiable uncertainties can be found, then the $\theta$ is said to be locally identifiable. The absence of unique parameter estimates for $\theta$ leads to practical non-identifiability. The practical identifiability of a parameter is hence contingent upon the nature, quality and quantity of data available to estimate the parameter as opposed to the structure and parameterization of the model.

	So, on the one hand, establishing the structural identifiability of parameters enables one to propose models that are not only appropriate representations of physical processes, but are also parameterized in such a way that the value of these parameters can be estimated from measurable data. On the other hand, establishing practical identifiability of parameters in any model helps design experiments that are minimal, informative and useful for parameter estimation.

	\subsection{A method to determine identifiability of kinetic models of metabolism}\label{sec:ident}
	We provide the mathematical framework for \textcolor{red}{identification} of parameters in kinetic models of metabolism in this section. A summary of the methodology in the form of a flow diagram is shown in Figure \ref{fig:ident-flowchart}. As indicated in Figure \ref{fig:ident-flowchart}a, the first step involves the construction of the kinetic model (Equation \ref{eq:kinstoich}) of the metabolic network with $n_r$ reaction fluxes.

@@

	%Nonetheless, all these explanations could only be verified due to the fact that the conditions for identifiability for both $v_1$ and $v_2$ are simple and readable due to the use of the Michaelis-Menten rate law to model these fluxes. Instead, in addition to presence of an allosteric interaction, when the convenience kinetic rate law is used to model this activation in $v_3$, we find that the complex expressions obtained for the three parameters preclude such an in-depth insight into the nature of persistently excitable experiments to identify  $v_3$. Thus, the only avenue through which experiments can be designed for the identification of parameters of $v_3$ would be through our proposed methodology wherein manual curation of experimental combinations required for identifiability are not necessary.

	%Based on the results we obtained in establishing the identifiability of parameters modeling all the fluxes of a small metabolic network, in the future we hope to extend this methodology to a larger metabolic network with more reactions.
	%Apart from determining experiments that are informative enough to identify parameters for each flux in a kinetic model, experimental design can also be used to design and develop a minimal set of experiments that informative enough to estimate all parameters in a metabolic network model. However, there is not only a scarcity of experimental design methods in literature devoted to the purpose of designing experiments for parameter identifiability in kinetic models of metabolism, the motivation to develop such methods is not well understood.

	\printbibliography

	%\clearpage
	%\section{Tables, Figures and Figure Captions:}

	%\begin{figure}[!tbhp]
	%	\centering{\includegraphics[width=1.0\textwidth,height=0.6\textheight, keepaspectratio]{figures/figure2/v1_V1max_v2_v5_2_exp_info}}
	%	\caption{The contribution of different experiments types used in a combination of two experiments ($j = \{1, 2\}$) that can practically identify parameter a) $V_1^{max}$, b) $K_1^{ac}$, c) $V_2^{max}$, d) $K_2^{pep}$, e) $V_e^{max}$ and f) $K_e^{fdp}(2)$. The percentages reflect the fractional contribution of each experiment type towards all identifiable data combinations.}\label{fig:exp_info}
	%\end{figure}

	%\clearpage

	%\section{Old Figures, Figure Captions and other miscellaneous text}	
	%Methods and tools for structural identification of parameters based on differential algebra \parencite{Ljung1994, Audoly2001, Bellu2007} and profile likelihood \parencite{Raue2014} are available.

\end{document}

>>>>

	Constraint-based models (CBMs) of metabolism are used to improve our understanding of metabolism by representing it as a stoichiometric network of reactions \parencite{Bordbar2014a}. The ability of CBMs to shine light on the nonintuitive interactions that govern cellular metabolism is leveraged to engineer and asses the impact of designs that alter the ability of a cell to grow, or produce a desired metabolite \parencite{Maia2016}. However, in CBMs, metabolism is assumed to operate under a pseudo steady state. Consequently, the metabolite concentrations within the metabolic network are assumed to be constant, and changes in metabolite concentrations are not modeled. Furthermore, since CBMs represent metabolism using only the stoichiometry of its constituent reactions, they do not account for the various non-catalytic regulatory interactions that are also responsible for metabolic function. These shortcomings prevent CBMs from being used to fully understand the steady state as well as the dynamic characteristics of metabolic networks.

	In contrast, the effects of regulatory interactions and changes in metabolite concentrations on different characteristics of metabolism can be studied using kinetic models of metabolism \parencite{Saa2017}. These models account for changes in metabolite concentrations subject to thermodynamic and regulatory constraints that underly metabolic networks in addition to their stoichiometry \parencite{Link2014}. Kinetic models can not only help us better understand lesser known and understood characteristics of metabolism like bistability \parencite{Kotte2014}, and their role in human health, but can also improve predictions about the impact of engineering design perturbations on metabolism, and help propose alternative designs to achieve metabolite production goals \parencite{Khodayari2016}.

	Kinetic models differ from CBMs in their use of mechanistic enzyme kinetics to model enzyme catalyzed fluxes within a metabolic network \parencite{Srinivasan2015,Saa2017}. The use of kinetic models requires information on the enzyme kinetic rate laws that will be used to model the fluxes within a metabolic network, as well as numerical values for the parameters used in these rate laws. Analyzing the ability of a metabolic network to exhibit dynamic characteristics like multiple steady states and oscillations, irrespective of the structure of the network, is one example where kinetic rate laws and parameter values play a crucial role \parencite{Srinivasan2017}.

	Despite their importance, the parameterization of kinetic models is still a problem for which solutions are a subject of debate within the modeling community. Typically, enzyme kinetic rate laws are parameterized based on in vitro observations of enzyme activity, as opposed to observations made under in vivo conditions \parencite{Heijnen2005,Smallbone2007}. However, some researchers have questioned their relevance for gleaning information on the dynamics of metabolism under in vivo conditions \parencite{Heijnen2005,Heijnen2013}. On the other hand, some reports have shown that despite the large uncertainties associated with parameters estimated based on in vivo experimental data \parencite{Link2014}, in vitro parameter estimates are a reasonable approximation of values that would be applicable under in vivo conditions \parencite{Davidi2016}.

	 Authors have sought to constrain and determine uncertainty in parameter estimates and associated model predictions by using Monte Carlo approaches for kinetic modeling of metabolism \parencite{Andreozzi2016a}. These approaches also allow for the integration of experimentally observed in vivo data. ORACLE \parencite{Wang2004} and Ensemble modeling \parencite{Tran2008}, are two such examples. These and other Monte Carlo kinetic modeling methods have been previously reviewed \parencite{Srinivasan2015}. 	 
	 Bayesian approaches to improve parameter estimation and quantify estimation uncertainty have also been proposed \parencite{Saa2016}. \textcite{Vanlier2013} provide a review of different Bayesian approaches to quantify parameter uncertainty.

	Related to the development of these methods to quantify parameter estimation uncertainty, the importance of model parameter identifiability, i.e., a necessary condition to estimate unique kinetic parameter values from experimental data, is often overlooked \parencite{Ljung1994,Berthoumieux2013}.

	Briefly, unique values can only be estimated for structurally identifiable parameters. If a parameter cannot be uniquely estimated due to its nonlinear relationships with other parameters (redundant model parameterization), or due to the inability to measure all possible metabolite concentrations (presence of unobservable system states), then the parameter is said to be structurally non-identifiable. Similarly, practical identifiability of a parameter concerns with the informativeness of the experimental data used for parameter estimation, and the ability to quantify any existing uncertainty in the parameter estimates. If the ability to estimate unique parameter values is compromised due to the inability of the available data to capture the requisite information needed to estimate the parameters in the modeled system, and the uncertainty in parameter estimates is not quantifiable, the parameter is said to be practically non-identifiable \parencite{Ljung1994,Rodriguez-Fernandez2006}.

	Authors have proposed ways to assess parameter identifiability by proposing approximate kinetic models of metabolism that utilize empirical enzyme kinetic rate laws with parameters that have physical significance, and are identifiable \parencite{Heijnen2005,Smallbone2007}. Significant work has also been done towards the development of methods for structural identification of parameters in kinetic models of metabolism \parencite{Ljung1994,Nikerel2009,Berthoumieux2013,Raue2014}. \textcite{Chis2011b} provide a comparison of the computational performance and implementation aspects of some of these methods.

	Methods to improve practical identifiability through a priori experimental design have also been developed, with a focus on kinetic models of metabolism \parencite{Gadkar2005a,Rodriguez-Fernandez2006,Vanlier2014a,Raue2014}. Some of these methods are limited by their applicability to approximate kinetic models only \parencite{Nikerel2009,Berthoumieux2013}, while some others suffer from computational limitations when applied to kinetic models of large metabolic networks \parencite{Gadkar2005a,Raue2014}.

	Most of the aforementioned methods assess identifiability (structural or practical) in the presence of dynamic time course concentration data. In this paper, we show that the identifiability of parameters for each individual reaction in kinetic models of metabolism can be done using just steady state fluxomics, metabolomics and proteomics data, and propose a methodology to do so. We also demonstrate how this method can also facilitate the design of experiments for generating the necessary steady state data to estimate unique parameter values for all reaction fluxes in a metabolic network. In doing so, we are able to propose the number and types of perturbations that will provide the most useful data for parameter estimation. We also assess the identifiability of different enzyme kinetic rate laws that are typically used to model fluxes in metabolic networks using steady state data. We illustrate our methodology using a small metabolic network model of glucoeneogenesis in \textit{Escherichia coli} \parencite{Kotte2014, Srinivasan2017} under the assumption that all intracellular metabolite concentrations and fluxes can be measured.

@@

	Parameter estimation methods based on optimization principles are used to determine parameter estimates from experimental data. Under the assumption that all intracellular metabolite concentrations and fluxes can be measured, a parameter estimation problem can be formulated as a nonlinear programming problem (Equation \ref{eq:lsqopt}) to estimate the values of enzyme kinetic parameters, $\theta \in \mathbb{R}^{n_p}$, based on the measured data. 
	\begin{subequations}\label{eq:lsqopt}
		\begin{align}
		\underset{\theta}{\mathrm{min}} &\sum_{k=1}^{n_x + n_r}\sum_{l=1}^{n_t}\left(\frac{y_{kl}^*-y_{kl}}{\sigma_{kl}^*}\right)^2\\
		&\theta_l \le \theta \le \theta_u
		\end{align}
	\end{subequations}
	Here $y \in \mathbb{R}^{n_x + n_r}$ is the combined vector of concentrations ($x\in \mathbb{R}^{n_x}$) and fluxes ($v \in \mathbb{R}^{n_r}$), at each time point $l={1, 2, ..., n_t}$. The minimization of the least squares error between the measured ($y^*$) and modeled ($y$) concentrations and fluxes, weighted by the variance in the experimental data $\sigma_{kl}^*$ for each concentration and flux, at each time point $l={1, 2, ..., n_t}$, is used as an objective function (Equation \ref{eq:lsqopt}a) for the optimization problem. The least squares parameter values are determined within fixed upper ($\theta_u$) and lower ($\theta_l$) bounds (Equation \ref{eq:lsqopt}b).

	Typically, not all metabolite concentrations and fluxes can be measured. Under these conditions, the formulation of the kinetic model and the parameter estimation problem shown in Equations (\ref{eq:kinstoich}) and (\ref{eq:lsqopt}) is different. These formulations are ubiquitously presented in other papers \parencite{Gadkar2005a,Rodriguez-Fernandez2006}. Hence, these will not be discussed in this paper.

@@

	The parameters in $\theta$ in any model (Equation \ref{eq:kinstoich}) are said to be structurally identifiable if, for an input-output mapping defined by $y = \Phi(\theta,u)$ for at least one input function $u$, any two values of parameters $\theta_1$ and $\theta_2$ satisfy the relationship in Equation (\ref{eq:stident}):
	\begin{align}\label{eq:stident}
	\Phi(\theta_1,u) = \Phi(\theta_2,u) \iff \theta_1 = \theta_2
	\end{align}
	Accordingly, if parameters in $\theta$ have a unique value, a finite number of non-unique values or an infinite number of values for all input functions, they are said to be structurally globally identifiable, locally identifiable or non-identifiable, respectively. So first, structural identifiability of parameters in a dynamic model helps establish the presence or absence correlations between different model parameters \parencite{Rodriguez-Fernandez2006}. Second, in instances where not all concentrations and fluxes can be measured, structural identification can also elucidate relationships between the unmeasurable and measured concentrations/fluxes. Consequently, the effect of model structure and parameterization on the ability to infer true parameter values from experimental data is determined by the structural identifiability of the parameter.

	Experimental data from many physical systems is usually noisy, and sometimes may not contain all requisite information pertaining to the system. When parameters are estimated on the basis of this data, the ability to estimate unique parameter values to satisfy Equation (\ref{eq:stident}) is referred to as practical identifiability. If a unique parameter value satisfying Equation (\ref{eq:stident}) can be estimated, then $\theta$ is said to be globally practically identifiable. Whereas, if parameter estimates with quantifiable uncertainties can be found, then $\theta$ is said to be locally identifiable. No unique parameter estimates can be found if $\theta$ is practical non-identifiable. The practical identifiability of a parameter is hence contingent upon the nature, quality and quantity of data available to estimate the parameter as opposed to the structure and parameterization of the model.

	Therefore, in conclusion, establishing the structural identifiability of parameters enables one to propose models that are not only appropriate representations of physical processes, but are also parameterized in such a way that the value of these parameters can be estimated from measurable data. On the other hand, establishing practical identifiability of parameters in any model helps design experiments that are minimal, informative and useful for parameter estimation.

	\subsection{A method to determine structural and practical identifiability of kinetic models of metabolism}\label{sec:ident}
	In this section, we provide the mathematical framework for \textcolor{red}{identification} of parameters in kinetic models of metabolism. A summary of the methodology in the form of a flow diagram is shown in Figure \ref{fig:ident-flowchart}. As indicated in Figure \ref{fig:ident-flowchart}a, the first step involves the construction of the kinetic model (Equation \ref{eq:kinstoich}) of the metabolic network with $n_r$ reaction fluxes.

@@

	\printbibliography	
\end{document}
